---
title: "Lake Champlain - metatranscriptomics"
author: "Sébastien Renaut"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document 
runtime: shiny
---

```{r setup, echo=F, message=F}
#install.packages("shiny")
library(shiny)
library(lubridate)
#library(taxize)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

***     

This is an R shiny app: it makes barplots and plots fold changes based on the Lake Champlain metatranscriptomics data.  
Currently, two variables can be adjusted: `Taxonomic level` and `cut-off`.      
Species-level count data per sample are first combined per `Taxonomic level` (species, genus, family, order, phylum)  
Then, only taxa above the a user specified `cut-off` value (in %) are plotted.   

***   

```{r prep_data, echo=F}

#don't run if don't need to...
if(file.exists("data/refseq.matrix.0.01")==F)
{

system("ls -1 /Users/jerry/Documents/CSBQ/shapiro/results/org_results/*RefSeq_annot_organism.tsv >data/RefSeq_org.tsv")
system("ls -1 /Users/jerry/Documents/CSBQ/shapiro/results/func_results/*RefSeq_annot_function.tsv >data/RefSeq_fun.tsv")
system("ls -1 /Users/jerry/Documents/CSBQ/shapiro/results/Subsystems_results/*reduced_modif >data/Subsys.tsv")

all.tsv = read.table("data/RefSeq_org.tsv", stringsAsFactors = F)
refseq.list = list()
all_spp = NULL
all.tsv.names = NULL
samples = NULL
locations = NULL
#first get all data into giant list
for(i in 1:nrow(all.tsv))
{
    refseq.list[[i]] = read.table(all.tsv[i,1],sep = "\t",stringsAsFactors = F)
    all_spp = c(all_spp,refseq.list[[i]][,3]) #keep all species
    replicate = substring(strsplit(all.tsv[i,1],split = "WatPhotz_")[[1]][2],4,4)
    location = substring(strsplit(all.tsv[i,1],split = "Champ")[[1]][2],1,3)
    date = strsplit(all.tsv[i,1],split = "-")[[1]][2]
    
    #get proper shortnames for the graph / samples
    samples = c(samples,paste(location,date,replicate,sep = "_"))
    locations = c(locations,location)
    }        
 
all_spp_unique = unique(sort(all_spp))

#then all data into giant data.frame    
refseq.matrix = data.frame(matrix(nrow = length(all_spp_unique),ncol = length(samples)))
rownames(refseq.matrix) = all_spp_unique
colnames(refseq.matrix) = samples

for(i in 1:nrow(refseq.matrix))
{
    for(j in 1:ncol(refseq.matrix))
        {
        temp = refseq.list[[j]][rownames(refseq.matrix)[i] == refseq.list[[j]][,3],1]
        if(length(temp) ==1) refseq.matrix[i,j] = temp
        }
        if(i %% 1000 == 0) print(paste("Running species ",i,", The time is: ",Sys.time(),sep = ""))
}

#keep only abundant ones (mean fraction above 0.01%...)
refseq.rowMeans = rowMeans(refseq.matrix,na.rm = T)
refseq.matrix.0.01 = refseq.matrix[refseq.rowMeans>0.01,]
dim(refseq.matrix.0.01)

#get taxonomy NOW
#add family,order, phylym with taxonomy...
refseq.taxonomy = data.frame(species = rownames(refseq.matrix.0.01),genus = NA,family = NA,order = NA,phylum = NA,stringsAsFactors = F)

for(j in 1:nrow(refseq.taxonomy))
{
    query = refseq.taxonomy$species[j]
    query = gsub(" sp.$","",query)
    if(query != "Bacteria") {
        taxonomy = tax_name(query = query, get = c("genus","family","order","phylum"), db = "ncbi",messages =F)
      #  if(!is.na(taxonomy$species)) taxonomy$species = tail(strsplit(taxonomy$species,split = " ")[[1]],1) #fix the species name problems...
        refseq.taxonomy[j,2:5] = taxonomy[1,3:6]
    }
    
    if(j %% 50 == 0) print(paste("Running sample ",j,"of ",nrow(refseq.taxonomy),", The time is: ",Sys.time(),sep = ""))
    
    Sys.sleep(1)
}

refseq.taxonomy[is.na(refseq.taxonomy[,2]),2] = "unknown"
refseq.taxonomy[is.na(refseq.taxonomy[,3]),3] = "unknown"
refseq.taxonomy[is.na(refseq.taxonomy[,4]),4] = "unknown"
refseq.taxonomy[is.na(refseq.taxonomy[,5]),5] = "unknown"

write.table(refseq.matrix.0.01,"data/refseq.matrix.0.01")
write.table(refseq.taxonomy,"data/refseq.taxonomy")
}
```  
  


```{r shiny_all, echo=F}
# Define UI for application that draws a histogram
ui <- fluidPage(

    # Application title
    titlePanel("Barplots"),

    # Sidebar with a slider input for number of bins 
    #sidebarLayout(
     #   sidebarPanel(
      fluidRow(
      column(6,
          # Input: Selector for variable to plot against mpg ----
            selectInput("taxo_level", label = "Taxonomic Level:",
            choices = c("species"="species","genus"="genus","family"="family","order"="order","phylum"="phylum"),selected = "genus")),
      
      column(6,    
          sliderInput("cutoff",
                        "Cut-off",
                        min = 0.1,
                        max = 5,
                        value = 1,step=0.1)
            )

        # Show a plot of the generated distribution
        ),
    plotOutput("distPlot")
    )


# Define server logic required to draw a histogram
server <- function(input, output) {

        output$distPlot <- renderPlot({
        
          #load data
          refseq.matrix.0.01 = read.table("data/refseq.matrix.0.01",stringsAsFactors = F)
          refseq.taxonomy = read.table("data/refseq.taxonomy",stringsAsFactors = F)
          
          #simplify by VARIABLE X (defined by user)
          variable = c(1:5)[colnames(refseq.taxonomy) == input$taxo_level]
          refseq.simplified =  refseq.matrix.0.01 %>% group_by(refseq.taxonomy[,variable]) %>% summarise_all(sum,na.rm = TRUE)
          
          #keep only the 1%
          refseq.simplified.rowMeans = rowMeans(refseq.simplified[,2:66],na.rm = T)
          refseq.gg = refseq.simplified[refseq.simplified.rowMeans>input$cutoff,]
          
          #construct ggplot2 object
          samples = colnames(refseq.matrix.0.01)
          locations = sapply(strsplit(colnames(refseq.matrix.0.01), "_"), "[[", 1)
          all_spp_m_ggplot = data.frame(unlist(refseq.gg[,2:66]))
          all_spp_m_ggplot[,2] = unlist(rep(refseq.gg[,1],65))
          all_spp_m_ggplot[,3] = c(t(matrix(rep(samples,  nrow(refseq.gg)),nrow=65,ncol=nrow(refseq.gg))))
          all_spp_m_ggplot[,3] = gsub("^....","",all_spp_m_ggplot[,3])
          all_spp_m_ggplot[,4] = c(t(matrix(rep(locations,nrow(refseq.gg)),nrow=65,ncol=nrow(refseq.gg))))
          all_spp_m_ggplot[,5] = factor(all_spp_m_ggplot[,4], levels=c('St1','St2','PRM'))
          colnames(all_spp_m_ggplot) = c("fraction","species","samples","locations","locations_f")
          
          #plot
          x = colorRampPalette(brewer.pal(12,"Paired"))
          p1=ggplot() + labs(title = paste("Lake Champlain - annotated",input$taxo_level,sep = " "),fill = "Taxonomy") +
            theme_bw() + 
            #scale_x_date(date_breaks = "months" , date_labels = "%b") +
            theme(legend.text = element_text(face="italic"),plot.title = element_text(hjust = 0.5, size=14, face="bold")) + scale_fill_manual(values = x(nrow(refseq.gg))) +
            geom_bar(aes(y = fraction, x = samples, fill = species),
                     data = all_spp_m_ggplot,stat="identity") + ylab("Percentage of annotated taxa")  + theme(axis.text.x = element_text(angle = 90, hjust = 1))
          
          #PDF (dimensions in inches)
          #dev.new()
          p1 + facet_grid(rows=vars(locations_f))
    })
}

# Run the application 
shinyApp(ui = ui, server = server,options=list(height = 650,width = "100%"))
```  
  
***   



```{r shiny_fc, echo=F}
# Define UI for application that draws a histogram
ui_fc <- fluidPage(

     # Application title
    titlePanel("Fold Changes"),

    # Sidebar with a slider input for number of bins 
    #sidebarLayout(
     #   sidebarPanel(
      fluidRow(
      column(6,
          # Input: Selector for variable to plot against mpg ----
            selectInput("taxo_level", label = "Taxonomic Level:",
            choices = c("species"="species","genus"="genus","family"="family","order"="order","phylum"="phylum"),selected = "genus")),
        
      column(6,  
          sliderInput("cutoff",
                        "Cut-off",
                        min = 0.1,
                        max = 5,
                        value = 1,step=0.1)
            )

        # Show a plot of the generated distribution
        ),
    plotOutput("distPlot")
    )

# Define server logic required to draw a histogram
server_fc <- function(input, output) {

        output$distPlot <- renderPlot({
        
          #load data
          refseq.matrix.0.01 = read.table("data/refseq.matrix.0.01",stringsAsFactors = F)
          refseq.taxonomy = read.table("data/refseq.taxonomy",stringsAsFactors = F)
          
          #simplify by VARIABLE X (defined by user)
          variable = c(1:5)[colnames(refseq.taxonomy) == input$taxo_level]
          refseq.simplified =  refseq.matrix.0.01 %>% group_by(refseq.taxonomy[,variable]) %>% summarise_all(sum,na.rm = TRUE)
          
          #keep only the 1%
          refseq.simplified.rowMeans = rowMeans(refseq.simplified[,2:66],na.rm = T)
          refseq.gg = refseq.simplified[refseq.simplified.rowMeans>input$cutoff,]
          
          #construct ggplot2 object
          samples = colnames(refseq.matrix.0.01)
          locations = sapply(strsplit(colnames(refseq.matrix.0.01), "_"), "[[", 1)
          all_spp_m_ggplot = data.frame(unlist(refseq.gg[,2:66]))
          all_spp_m_ggplot[,2] = unlist(rep(refseq.gg[,1],65))
          all_spp_m_ggplot[,3] = c(t(matrix(rep(samples,  nrow(refseq.gg)),nrow=65,ncol=nrow(refseq.gg))))
          all_spp_m_ggplot[,3] = gsub("^....","",all_spp_m_ggplot[,3])
          all_spp_m_ggplot[,4] = c(t(matrix(rep(locations,nrow(refseq.gg)),nrow=65,ncol=nrow(refseq.gg))))
          all_spp_m_ggplot[,5] = factor(all_spp_m_ggplot[,4], levels=c('St1','St2','PRM'))
          colnames(all_spp_m_ggplot) = c("fraction","species","samples","locations","locations_f")

          ##Fold changes
all_spp_m_ggplot$dates = unlist(strsplit(all_spp_m_ggplot$samples,"_"))[seq(1,(nrow(all_spp_m_ggplot)*2),by = 2)]
all_spp_m_ggplot$dates = ymd(all_spp_m_ggplot$dates)

#Keep only fractions
fraction = as.data.frame(all_spp_m_ggplot$fraction)

#dplyr fractions data per replicate, per location and per species
all_spp_m_ggplot.replicates_merged = fraction %>% group_by(all_spp_m_ggplot$locations,all_spp_m_ggplot$dates,all_spp_m_ggplot$species) %>% summarise_all(mean)

#add columns for FoldChanges.
all_spp_m_ggplot.replicates_merged$fold_change_mean = 0
all_spp_m_ggplot.replicates_merged$fold_change = 0

#add the locations as factors as well + give column names
all_spp_m_ggplot.replicates_merged$locations_f = 0
colnames(all_spp_m_ggplot.replicates_merged) = c("locations","dates","species","fraction","fold_change_mean","fold_change","locations_f")
all_spp_m_ggplot.replicates_merged = as.data.frame(all_spp_m_ggplot.replicates_merged)
all_spp_m_ggplot.replicates_merged$locations_f = factor(all_spp_m_ggplot.replicates_merged[,1], levels=c('St1','St2','PRM'))

####Fold change progression from Time 0 (logFC = 1)
for(i in 1: nrow(all_spp_m_ggplot.replicates_merged))
{
  #per species
  temp = all_spp_m_ggplot.replicates_merged[all_spp_m_ggplot.replicates_merged$species == all_spp_m_ggplot.replicates_merged[i,3],]
  
  #per location
  temp = as.data.frame(temp)
  temp2 = temp[temp$locations == all_spp_m_ggplot.replicates_merged[i,1],]
  
  #value divided by mean value
  all_spp_m_ggplot.replicates_merged$fold_change_mean[i] = log(all_spp_m_ggplot.replicates_merged[i,4] / mean(temp2[,4], na.rm =T),10)
  #value divided by Time 0
  all_spp_m_ggplot.replicates_merged$fold_change[i] = log(all_spp_m_ggplot.replicates_merged[i,4] / temp2[temp2[,2]=="2016-06-01",4],10)
    }

#ggplot object
x = colorRampPalette(brewer.pal(12,"Paired"))
p2=ggplot() + labs(title = "Lake Champlain - annotated genera",fill = "Taxonomy") +
  theme_bw() + 
  theme(legend.text = element_text(face="italic"),plot.title = element_text(hjust = 0.5, size=14, face="bold")) + scale_colour_manual(aesthetics = "colour",values = x(  length(unique(sort(all_spp_m_ggplot.replicates_merged[,3])))    )) +
  geom_line(aes(y = fold_change, x = dates, colour = species,group=species),size = 2,data = all_spp_m_ggplot.replicates_merged,stat="identity") +
  ylab("Fold change (log) with respect to Time 0 (June 1st)")  + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(colour=input$taxo_level) +
  scale_x_date(date_breaks = "months" , date_labels = "%b")

#PDF (dimensions in inches)
p2 + facet_grid(rows=vars(locations_f))
    })
}

# Run the application 
shinyApp(ui = ui_fc, server = server_fc,options=list(height = 550,width = "100%"))
``` 

  
***  
  